---
title: "Exercise 0"
output: html_notebook
---

# Exercise 0.1

We will work with the iris dataset, which contains size measurements of different species of flowers. This dataset is built into R and can be loaded with the simple command below.

```{r}
data('iris')
```

Let's take a look at the data.

```{r}
head(iris)
```

Read about these functions, try them on `iris`, and try to understand their output:

* `dim`
* `length`
* `names`
* `summary`

# Exercise 0.2

Say we want to draw random samples from `iris`. How do we do this? Go to Google and search for "draw random samples from data frame R". One of the search results will be the following question on **StackOverflow**:

> https://stackoverflow.com/questions/8273313/sample-random-rows-in-dataframe

Note how different the solutions are. Try loading the `dplyr` library and using the `sample_n()` function to draw random samples from `iris`.


# Exercise 0.3

In this exercise we will try to write a script.

* Make a directory structure to store scripts for the exercises: r_for_statistics > exercises
* Make a script in the exercises directory: exercise_0_2.R
* Load the “ggplot2” and “magrittr” libraries
    * If you didn’t already install these, do it now
* Write the following code:
    * ggplot(iris, aes(Sepal.Width)) + geom_histogram(bins=10)
* Try evaluating:
    * A single line of code (ctrl + enter)
    * A block of code (highlight + ctrl + enter)
    * The entire file (ctrl + alt + enter)
* Write comments for the code





