---
title: "Exercise 1"
output: html_notebook
---

# Exercise 1.1

Let's load the iris dataset again.

```{r}
data("iris")
```

Remember that there often many ways to solve the same problem. The three lines of code below produce the same result in three different ways.

```{r}
sum(iris$Sepal.Length > 5)
iris %>% summarize(sum(Sepal.Length > 5)) %>% pull()
iris %>% filter(Sepal.Length > 5) %>% nrow()
```

Try making a histogram of the `Sepal.Width` parameter again and change the **label on the x-axis**. Go to the `ggplot2` reference site (https://ggplot2.tidyverse.org/reference/index.html), find the reference for `geom_histogram()`, and figure out how to change the x-axis label.

When you've done this, give your plot a title.

# Exercise 1.2

Make a scatterplot of two variables in the iris dataset. If you don't know what ggplot function to use for this, try searching in Google.

When you've done this, try coloring the points in the scatterplot by the `Species` variable.

# Exercise 1.3

Make a new variable in `iris` that is a the difference between the `Sepal.Width` and `Petal.Length` variables, and call it `Length.Diff`. Hint: use the `mutate()` function.

# Exercise 1.4

Try making a linear regression between two of the measurement variables using `lm()`. Try experimenting the model, adding more variables and interactions between variables.






